---
title: 'Tidyverse Overview'
output:
  html_document:
    code_download: yes
    fontsize: 8pt
    highlight: textmate
    number_sections: no
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(fig.width=5.25)
knitr::opts_chunk$set(fig.height=4.0)
knitr::opts_chunk$set(fig.align='center') 
knitr::opts_chunk$set(results='hold') 
```



# Tidyverse 
- Family of packages developed to streamline graphing, data manipulation, data wrangling, and programming
- Main packages of interest for us today: ggplot2, dplyr, tidyr (these have evolved to the point that they have replaced packages we converted in the r class: reshape2 and plyr)
- These packages can work on a modern type of data frame called a tibble 
- ggplot2 is for graphing
- dplyr is for data manipulation (think excel pivot tables on steroids)
- tidyr is for data wrangling (long to wide format and back again)
- These packages can use piping (in dplyr which calls purrr) which makes it easier to string together commands 


# Wrangling 
- tidyr makes data tidy
- 2 workhouse functions
    - gather: Makes wide data long
    - spread: Makes long data wide
- 2 supplemental functions
    - separate/unite: breaks/merges columns into multiple/one column
    - extract: takes groups within a column to make new columns 

## Example Dataset
- Longitudinal data of trust over 5-time points (game plays) with the within-subject variable of vision
- Do you trust your partner in the game of "trouble" when you can or cannot see them 
- Covariate (how "judging" personality type they are 1-3 scale)
- Data is in typical wide SPSS-like format
- [Download Data](www.alexanderdemos.org/Mixed/TrustWideData.csv)

```{r}
setwd('/Users/rachel/Box\ Sync/R\ Coursework/Class18/')
TrustWide<-read.csv("TrustWideData.csv")
```

```{r, echo=FALSE}
library(knitr); library(kableExtra) 
kable(head(TrustWide), "html", booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```


- This is not so useful for taking means and doing stats, so we will convert it to long format

## Convert to Long
- Process in words: 
    - We will use piping %>% to help us **pass** the dataset along to each function (so you ignore when the function asks for dataframe first)
    - gather(new variable name, new value name, variables to merge)

```{r}
library(tidyr)
library(dplyr)
TrustLong<-TrustWide %>% gather(Condition, TrustFeeling,NoSee_0:See_4)
str(TrustLong)
```


```{r, results='asis',echo=FALSE}
kable(head(TrustLong), "html", booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```


- Personality applies to the subject, so every time we see the subject names, we need their Personality score
- Our two variables (Vision & Time) are all stuck together. Good for SPSS RM anova, but not helpful for us.
    - separate(variable to split, c("what do we call them"), what separates them, convert them to back to numeric/integers if they are numbers) 

```{r}
TrustLong.Final<-TrustLong %>% separate(Condition,c("Vision","Time"),sep="_", convert = TRUE)
```

```{r, results='asis',echo=FALSE}
kable(head(TrustLong.Final), "html", booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

- You can do this all at once!

```{r}
TrustLong.Final<-
  TrustWide %>% 
  gather(Condition, TrustFeeling,NoSee_0:See_4) %>% 
  separate(Condition,c("Vision","Time"),sep="_", convert = TRUE)
```

### Convert back to wide
- You just have to inverse the process. First, unite the variables into one column and then spread them out

```{r}
TrustWide.Again<-
  TrustLong.Final %>% 
  unite(Condition,c("Vision","Time"),sep="_") %>% 
  spread(Condition, TrustFeeling)
```

```{r, results='asis',echo=FALSE}
kable(head(TrustWide.Again), "html", booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

# Data Manipulation 
- dplyr lets you calculate means, sd or any other statistic you may want on the data based on how the data was wrangled.
- 3 useful functions
  - group_by (how to cut data)
  - filter (subset on the fly)
  - summarise (what descriptive to conduct)
  - do (do more complex stats) [with the help of broom package which converts regressions to tibbles]
  - mutate (add stats back to data frame on the fly)
  
## Group and summarise
- We can calculate means and sd per group
```{r}
Means<-TrustLong.Final %>%
  group_by(Vision,Time) %>%
  summarise(MeanTrust=mean(TrustFeeling),
            SDTrust=sd(TrustFeeling))
```

```{r, results='asis',echo=FALSE}
kable(Means, "html", booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

- we can also calculate a correlation per subject separately per vision condition

```{r}
library(broom)
CorrResult<-TrustLong.Final %>%
  group_by(Vision,Subject) %>%
  do(tidy(cor.test(.$Time, .$TrustFeeling)))
```

```{r, results='asis',echo=FALSE}
kable(head(CorrResult), "html", booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

- or we can also run regression per subject separately per vision condition 

```{r}
RegresResult<-TrustLong.Final %>%
  group_by(Vision,Subject) %>%
  do(tidy(lm(TrustFeeling ~ Time, data=.)))

```

```{r, results='asis',echo=FALSE}
kable(head(RegresResult), "html", booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

- You mutate the dataset you just created and add a column created by another column

```{r}
RegresResult <-
  RegresResult %>%
  mutate(Sig = if_else(p.value < .05,1,0))
```

```{r, results='asis',echo=FALSE}
kable(head(RegresResult), "html", booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

- Now you can take that new dataset and summarize it (mean slope estimates just for Time by significant vs nonsignificant results) [*Note: dplyr::filter is used because its conflicts with function in another package. Thus this forces filter function to come from the dplyr package.*]

```{r}
RSum<-RegresResult  %>%
  group_by(Vision,Sig) %>%
  dplyr::filter(term=="Time") %>%
  summarise(N=length(Sig),
            MeanSlope=mean(estimate))
```

```{r, results='asis',echo=FALSE}
kable(RSum, "html", booktabs = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

- Note: You can do this all at once because its conduct in the order you call it

```{r}
RegresFinal<-TrustLong.Final %>%
  group_by(Vision,Subject) %>%
  do(tidy(lm(TrustFeeling ~ Time, data=.))) %>%
  mutate(Sig = if_else(p.value < .05,1,0)) %>%
  group_by(Vision,Sig) %>%
  dplyr::filter(term=="Time") %>%
  summarise(N=length(Sig),
            MeanSlope=mean(estimate))
```

- You can pipe all these results into a ggplot

# ggplot
Inspired by "The Grammar of Graphics" Leland Wilkinson 1999 

> "Destined to become a landmark in statistical graphics, this book provides a formal description of graphics, particularly static graphics, playing much the same role for graphics as probability theory played for statistics."Journal of the American Statistical Association Former VP at SPSS Inc. Founder of SYSTAT. Adjunct Professor of Statistics at Northwestern University. He is also affiliated with the Computer Science department at The University of Illinois at Chicago.

- Manual: http://ggplot2.tidyverse.org/reference/

## Grammar 
- Data (*data = *): **Data.Frame** to be mapped 
- Aesthetic mapping (*aes( )*): *x = *, *y = *, *group =*  from your data. Can also add additional mapping into the aes:
    - (*color = *, *shape = *, *size = *, *fill = *, *alpha =*, etc): These can be fixed values are variables
        - Note: if place these calls into aes(x=IV, y=DV, color=subjectID), then the subject will appear in your legend with those colors. You can move some these maps into geom call to avoid that
- Geometric object (*geom_*): *bar*, *point*, *line*, *ribbons*, shapes you want to graph based on your *as* 
    - Position adjustments (*position = *): goes with *geom_* call such as *position_dodge* (don't overlap), *position_identity* (leave as read in), *position_jitter*  (jitters data points in scatterplot)
- Statistical transformations (*stats_*): On the fly transforms (such as averaging): can be used instead of geoms (*Note*: I prefer to calculate stats outside of the plot when possible as its easier to see what you are doing)
- Coordinate system (*coord_*): do you want to add *coord_cartesian()*, *coord_polar()*, etc
- Scales (*scale_*) or simply (xlim, ylim).  Override defaults to control many aspects of the graphs
- Faceting (*facet_*): visual subsets: two options *grid* or *wrap*

### Layering
- First, you put on your jacket; then you put on your shoes, next underwear, finally you shower, right?
- ggplot has very specific order that you should generally follow. 
    - what is my data, what are my mappings, what are geoms, how should I position them, and last what do I want to do with the look 
- The order of which you add calls is the order they appear. So later calls will override earlier calls

