---
title: "RachelSchroeder_HMWK2"
output: html_document
---
Read in the data
```{r, echo=FALSE}
setwd('/Users/rachel/Box\ Sync/R\ Coursework/HW2/')
hyperactivity_study<-read.csv('OneWayData.csv')

hyperactivity_study
```

Check the means and SS calculations with dpylr
```{r}
library(dplyr)
Means.Table<-hyperactivity_study %>%
  group_by(Condition) %>%
  summarise(N=n(),
            Means=mean(Hyperactivity),
            SS=sum((Hyperactivity-Means)^2),
            SD=sd(Hyperactivity),
            SEM=SD/N^.5)

knitr::kable(Means.Table, digits = 2)
```

Run the ANOVA
```{r}
library(afex)

ANOVA.hyperactivity<-aov_car(Hyperactivity~Condition + Error(Subject), 
                  data=hyperactivity_study)
ANOVA.hyperactivity
```

```{r, fig.height=3, fig.width=7.5} 
#Test for normality

library(HH)
hov(Hyperactivity~Condition, data=hyperactivity_study)
hovPlot(Hyperactivity~Condition, data=hyperactivity_study) 
```

```{r, fig.height=3, fig.width=3}
qqnorm(hyperactivity_study$Hyperactivity)
qqline(hyperactivity_study$Hyperactivity)
```
Data are normal and homogeneous. The effect of the condition is sig at 0.001; need to do follow up tests to see which are different from each other

They hypothesized: 
(a) that children who experience a psychological trauma will display symptoms that are similar to children who suffer from attention-deficit/hyperactivity disorder (ADHD) and 
(b) that the ADHD and trauma group scores would be higher in symptoms of hyperactivity than those of the control group

Ha: Trauma group scores = ADHD group scores on hyperactivity
Hb: Trama + ADHD combined group hyperactivity scores = Control group scores

```{r}
#confirm the order of the conditions
Fitted.Model<-emmeans(ANOVA.hyperactivity, ~Condition)
Fitted.Model
```

Plot the ANOVA
```{r, fig.height=2, fig.width=2}
library(ggplot2)

Means.Table$Condition<-factor(Means.Table$Condition, 
                              levels = c("ADHD","Control","Trauma"),
                              labels = c("ADHD","Control","Trauma"))

Plot.1<-ggplot(Means.Table, aes(x = Condition, y = Means))+
  geom_col(aes(fill=Condition))+
  scale_y_continuous(expand = c(0, 0))+ # Forces plot to start at zero
  geom_errorbar(aes(ymax = Means + SEM, ymin= Means - SEM), 
                position=position_dodge(width=0.9), width=0.25)+
  xlab('Mental Health Status')+
  ylab('BASC Hyperactivity Scores')+
  theme_bw()+
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line(),
        legend.title=element_blank())
Plot.1
```

Using linear contrasts so that I can reduce the number of tests to 2
Order: ADHD, Control, Trauma
Ha: -1, 0, 1
Hb: 1, -2, 1

```{r}
Test1 <- list(
  Ha = c(-1, 0, 1))

contrast(Fitted.Model, Test1, adjust = "none")
```
Ha: p=0.0532, supporting the hypothesis that kids with ADHD and Trauma are not significantly different in hyperactive behavior


```{r}
Test2 <- list(
  Hb = c(1, -2, 1))

contrast(Fitted.Model, Test2, adjust = "none")
```
Hb: p<.0001, Trama + ADHD combined group hyperactivity scores are significantly different from Control group scores

###########################################
FOLLOW UP: Let's make sure that the ADHD and Trauma scores are sig different from the Control
Hc: -1, 1, 0 #ADHD
Hd: 0, 1, -1 #Trama
Will add correction since these are extra tests
```{r}
Test3 <- list(
  Hc = c(-1, 1, 0))

contrast(Fitted.Model, Test3, adjust = "sidak")
```
Hc: p<.0001, ADHD is sig dif from Control


```{r}
Test4 <- list(
  Hd = c(0, 1, -1))

contrast(Fitted.Model, Test4, adjust = "sidak")
```
Hd: p<.0001, Trauma is sig dif from Control

Calculate Effect Size
```{r}
library(effsize)
## Run Effect size Ha
hyperactivity_study.Ha <- hyperactivity_study %>% filter(Condition== "ADHD" | 
                                  Condition== "Trauma" ) %>% 
  droplevels() 
D1<-cohen.d(Hyperactivity~Condition,data=hyperactivity_study.Ha)

## Run Effect size Hb
hyperactivity_study$Hbtest<-with(hyperactivity_study,
     ifelse(Condition== "Control" | Condition== "No", "ADHD", "Trauma"))
# Means
library(dplyr)
Means.Hb<-hyperactivity_study %>%
  group_by(Hbtest) %>%
  summarise(N=n(),
            Means=mean(Hyperactivity),
            SD=sd(Hyperactivity))

D2<-cohen.d(Hyperactivity~Hbtest,data=hyperactivity_study)

## Run Effect size Hc
hyperactivity_study.Hc <- hyperactivity_study %>% filter(Condition== "ADHD" | 
                                  Condition== "Control" ) %>% 
  droplevels() 
D3<-cohen.d(Hyperactivity~Condition,data=hyperactivity_study.Hc)

## Run Effect size Hd
hyperactivity_study.Hd <- hyperactivity_study %>% filter(Condition== "Trauma" | 
                                  Condition== "Control" ) %>% 
  droplevels() 
D4<-cohen.d(Hyperactivity~Condition,data=hyperactivity_study.Hd)

D1
D2
D3
D4
```

